### 2014-07-28 Studies/2012_Chicago/Report/Chicago_Fire.tex
Chicago: Address some edits from reviewer

\headerB{
Simulation of an Attic Fire \\
in a Wood Frame Residential Structure \\
Chicago, IL \\
}

\headerC{
\flushright{
Craig G. Weinschenk \\
Daniel Madrzykowski \\
Kristopher J. Overholt \\
{\em Fire Research Division \\
Engineering Laboratory \\
Gaithersburg, Maryland, USA} \\ }
}

\flushright{\today \\
}

@@

\flushright{
\includegraphics[width=2.in]{../../../Bibliography/nistident_flright_vec} \\[.3in]
}

@@

@@

\chapter{Introduction}
Part of the mission of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Cook_County,Grosshandler:Station,Bryner:Charleston,barowy:texas}. The objective of all of these studies has been to improve the safety of fire fighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago firefighter. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using Fire Dynamics Simulator and Smokeview to provide insight into the fire development and thermal conditions that might have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident follows the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding times are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the Sector 3 door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a horseshoe load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the back door to the covered porch and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door to the first floor apartment, preceded to walk in, saw no fire then backed out.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~(\ref{fig:simp_geom}).
\begin{table}
\centering
\captionof{table}{Abridged NISOH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &                                        \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on 2nd floor with 1\sfrac{3}{4}-inch hoseline. \\[.25cm]
17:21:30 & 1st floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~(\ref{fig:simp_geom}). This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~(\ref{fig:simp_geom})), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~(\ref{fig:simp_geom})) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~(\ref{fig:simp_geom})). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~(\ref{fig:simp_geom}). The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~(\ref{fig:alpha_ex}) shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}), respectively.

\begin{figure}[h!]
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
\includegraphics[width=\textwidth]{../Figures/1st_Floor_Metric} \\
\includegraphics[width=\textwidth]{../Figures/2nd_Floor_Metric}
\end{tabular*}
\caption[Plan view of 1st (top) and 2nd (bottom) floor of structure.]{Plan view of 1st (top) and 2nd (bottom) floor of structure. Stairs and key rooms are identified using measurements collected by NIST.}
\label{fig:simp_geom}
\end{figure}

\begin{figure}[h!]
\centering
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
   \scalebox{1.0}{ \includegraphics[width=3in]{../Figures/exterior_alpha} } &
   \scalebox{1.0}{ \includegraphics[width=2.67in]{../Figures/exterior_charlie} } \\
\end{tabular*}
\caption{Photographs showing the front (left) and rear (right) exterior of the structure after the incident.}
\label{fig:alpha_ex}
\end{figure}
%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%\begin{figure}[h!]
%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%\end{tabular*}
%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%\label{fig:smv_interior}
%\end{figure}

@@

\begin{table}
\centering
\captionof{table}{Timeline of Ventilation Changes in the Simulation}\label{tab:vents}
\begin{tabular}{cccll}
\toprule[1.5pt]
Baseline    &  Alternative  &  Ventilation  &  Type         &  Side of         \\
Simulation  &  Simulation   &  Area         &               &  Structure       \\
{[s]}       &  {[s]}        &  [m$^2$]      &               &                  \\
\midrule
0           &  0            &  0.11         &  Window       &  Rear            \\
40          &  40           &  0.08         &  Breach (x2)  &  Front and Rear  \\
45          &  45           &  0.12         &  Breach       &  Left            \\
55          &  55           &  0.12         &  Breach       &  Rear            \\
130         &  N/A          &  1.41         &  Door         &  Rear            \\
135         &  N/A          &  0.2          &  Breach       &  Left            \\
160         &  160          &  0.80         &  Door         &  Interior        \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=.60\textwidth]{../Figures/smv_exterior_grid}
\caption{Front side of the structure with 10 cm computational mesh.}
\label{fig:geom_grid}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=.60\textwidth]{../Figures/smv_exterior_mesh}
\caption[Front-right side of the structure within the computational domain (with multiple meshes).]{Front-right side of the structure within the computational domain (with multiple meshes). Each of the 16 meshes is 3~m (9.8~ft) by 10~m (33~ft) by 5~m (16.4~ft).}
\label{fig:mult_mesh}
\end{figure}

\begin{table}
\centering
\captionof{table}{Relevant Modeling Input Parameters}\label{tab:mod_param}
\begin{tabular}{lll}
\toprule[1.5pt]
Parameter & Description & Reference  \\
\midrule
 Simulation Time                   &  5 min                                                        & -- \\[.25cm]
 Grid Cell Size                    &  10 cm                                                        & Section~\ref{mesh} \\ [.25cm]
 Ambient Temperature               &  25 $^{\circ}$C                                               & -- \\ [.1cm]
\multirow{4}{*}{Reaction: Wood}    &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$                     & \multirow{4}{*}{Section~\ref{fire}} \\
                                   &  CO Yield: 0.004 \; ${\mathrm{kg_{CO}}/\mathrm{kg_{wood}}}$   & \\
                                   &  Soot Yield: 0.015 \; ${\mathrm{kg_{C}}/\mathrm{kg_{wood}}}$  & \\
                                   &  Heat of Combustion: 16,400~kJ/kg                             & \\ [.25cm]
\multirow{2}{*}{Peak HRR}          &  Attic Fire: 9~MW                                             & \multirow{2}{*}{Section~\ref{fire}}\\ 
                                   &  Porch Fire: 3.3~MW                                           & \\ [.25cm]                     
\multirow{3}{*}{Material: Wood}    &  k: 0.12 \; \si{W/(m.K)})                                     & \multirow{3}{*}{Section~\ref{matl}} \\
                                   &  $\rho$: 510 \; kg/m$^3$                                      & \\
                                   &  c$_{p}$: 1.38 \; \si{kJ/(kg.K)}                              & \\[.25cm]
\multirow{3}{*}{Material: Gypsum}  &  k: 0.17 \; \si{W/(m.K)}                                      &  \multirow{3}{*}{Section~\ref{matl}}\\ 
                                   &  $\rho$: 930 \; kg/m$^3$                                      & \\ 
                                   &  c$_{p}$: 1.09 \; \si{kJ/(kg.K)}                              & \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=.80\textwidth]{../Figures/Chicago_Fire_HRR}
\caption[HRR versus time from FDS simulations of the West 50th St. fire.]{HRR versus time from FDS simulations of the West 50th St. fire. The first vertical line indicates when the back door failed, and the second vertical line indicates when the interior hallway door failed.}
\label{fig:hrr}
\end{figure}

In Fig.~(\ref{fig:hrr}), the solid line represents the baseline FDS simulation (this represents the on-scene conditions). The dashed line represents an alternative series of events in which the rear door does not get opened and the subsequent breach on the left-side of the structure does not occur. The dash-dotted line represents the prescribed (or nominal) HRR that was input into FDS. As shown in Fig.~(\ref{fig:hrr}), the HRR output from FDS is less than the prescribed HRR because the fire in the structure becomes ventilation limited. The maximum HRR value is slightly less than 9~MW before the fire becomes ventilation limited and begins to decay. The HRR is reduced to 5~MW as the oxygen inside the structure is consumed. There is an insufficient amount of oxygen in the structure to combust all of the fuel, therefore combustion occurs locally at the vents where fresh air is available. Figure~(\ref{fig:smv_ext_fire}) shows flaming combustion through the failed window on the rear side of the structure and attic breach just after the second fire ignites at 40~s. After additional wall breaches occur at 45~s and 55~s (see Table~\ref{tab:vents}), the abrupt change in ventilation causes the HRR to increase to 9~MW before it decreases to approximately 6~MW. The HRR then increases towards a sustainable steady-state value.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=.675\textwidth]{../Figures/smv_exterior_fire}
\caption{Smokeview rendering of exterior combustion occurring on the rear side of the structure.}
\label{fig:smv_ext_fire}
\end{figure}

@@

As the fire grows in the attic and rear porch, the pressure rises. Figure~(\ref{fig:pres_159s}) shows the pressure conditions in the structure just before the interior door failed in the simulation. On the rear porch, the pressure rise ranges from a 4~Pa over pressure 1~m off the floor to greater than 10~Pa at the ceiling. Gases at an elevated pressure will always move towards a region of lower pressure. The door serves as a boundary between the high and low pressure regions. Once the door fails open, the high pressure gases flow towards the low pressure regions in the house, hallway, kitchen, and front stairwell. The velocity vectors shown in Fig.~(\ref{fig:velo_175s}), 15~s after the interior door fails, confirm this flow direction within the structure.

>>>>

\begin{flushright}
\fontsize{20}{24}\selectfont
\bf{NIST Technical Note XXXX}
\end{flushright}

\headerB{
Simulation of an Attic Fire \\
in a Wood Frame Residential \\
Structure - Chicago, IL \\
}

\normalsize

\headerC{
{
\flushright{
Craig G. Weinschenk \\
Daniel Madrzykowski \\
Kristopher J. Overholt \\

\vspace*{2\baselineskip}

\begingroup
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.XXXX}{http://dx.doi.org/10.6028/NIST.TN.XXXX}
\endgroup
}

@@

\flushright{

\includegraphics[width=2.in]{../../../Bibliography/nistident_flright_vec} \\[.3in]
}
}
}

\pagenumbering{roman}

\begin{flushright}
\fontsize{20}{24}\selectfont
\bf{NIST Technical Note XXXX}
\end{flushright}

\headerB{
Simulation of an Attic Fire \\
in a Wood Frame Residential \\
Structure - Chicago, IL \\
}

\headerC{
\flushright{
Craig G. Weinschenk \\
Daniel Madrzykowski \\
Kristopher J. Overholt \\
{\em Fire Research Division \\
Engineering Laboratory} \\

\vspace*{2\baselineskip}

\begingroup
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.XXXX}{http://dx.doi.org/10.6028/NIST.TN.XXXX} \\
\endgroup

\vspace*{2\baselineskip}

August 2014}}

\flushright{\includegraphics[width=1in]{../../../Bibliography/doc} }

@@

\flushright{Certain commercial entities, equipment, or materials may be identified in this \\
document in order to describe an experimental procedure or concept adequately. \\
Such identification is not intended to imply recommendation or endorsement by the \\
National Institute of Standards and Technology, nor is it intended to imply that the \\
entities, materials, or equipment are necessarily the best available for the purpose. \\
}

\vspace{3in}

\large
\flushright{\bf National Institute of Standards and Technology Technical Note XXXX \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~XX, \pageref{LastPage} pages (August 2014) \\
% http://dx.doi.org/10.6028/NIST.TN.XXXX \\
CODEN: NTNOEF }

\hspace{1in}

@@

\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of a fire captain. The inputs for the FDS simulations are documented in this report.  The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas.  The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door.  After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at face height.  Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. The captain was overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure but later succumbed to his injuries.

\chapter{Introduction}
Part of the mission of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of all of these studies has been to improve the safety of fire fighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago fire captain. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using Fire Dynamics Simulator and Smokeview to provide insight into the fire development and thermal conditions that might have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident follows the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding times are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the [interior door opening into the covered porch on the rear of the house] and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door to the first floor apartment, preceded to walk in, saw no fire then backed out.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~(\ref{fig:simp_geom}).
\begin{table}
\centering
\captionof{table}{Abridged NOISH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &                                        \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on 2nd floor with 1\sfrac{3}{4}-inch hoseline. \\[.25cm]
17:21:30 & 1st floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~(\ref{fig:simp_geom}). This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~(\ref{fig:simp_geom})), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~(\ref{fig:simp_geom})) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~(\ref{fig:simp_geom})). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~(\ref{fig:simp_geom}). The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~(\ref{fig:alpha_ex}) shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}), respectively.

\begin{figure}[!ht]
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
\includegraphics[width=\textwidth]{../Figures/1st_Floor_Metric} \\
\includegraphics[width=\textwidth]{../Figures/2nd_Floor_Metric}
\end{tabular*}
\caption[Plan view of 1st (top) and 2nd (bottom) floor of structure.]{Plan view of 1st (top) and 2nd (bottom) floor of structure. Stairs and key rooms are identified using measurements collected by NIST.}
\label{fig:simp_geom}
\end{figure}

\begin{figure}[!ht]
\centering
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
   \scalebox{1.0}{ \includegraphics[width=3in]{../Figures/exterior_alpha} } &
   \scalebox{1.0}{ \includegraphics[width=2.67in]{../Figures/exterior_charlie} } \\
\end{tabular*}
\caption{Photographs showing the front (left) and rear (right) exterior of the structure after the incident.}
\label{fig:alpha_ex}
\end{figure}
%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%\begin{figure}[!ht]
%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%\end{tabular*}
%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%\label{fig:smv_interior}
%\end{figure}

@@

\begin{table}
\centering
\captionof{table}{Timeline of ventilation changes in the simulation}\label{tab:vents}
\begin{tabular}{cccll}
\toprule[1.5pt]
Baseline    &  Alternative  &  Ventilation  &  Type         &  Side of         \\
Simulation  &  Simulation   &  Area         &               &  Structure       \\
{[s]}       &  {[s]}        &  [m$^2$]      &               &                  \\
\midrule
0           &  0            &  0.11         &  Window       &  Rear            \\
40          &  40           &  0.08         &  Breach (x2)  &  Front and Rear  \\
45          &  45           &  0.12         &  Breach       &  Left            \\
55          &  55           &  0.12         &  Breach       &  Rear            \\
130         &  N/A          &  1.41         &  Door         &  Rear            \\
135         &  N/A          &  0.2          &  Breach       &  Left            \\
160         &  160          &  0.80         &  Door         &  Interior        \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=.60\textwidth]{../Figures/smv_exterior_grid}
\caption{Front side of the structure with 10 cm computational mesh.}
\label{fig:geom_grid}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=.60\textwidth]{../Figures/smv_exterior_mesh}
\caption[Front-right side of the structure within the computational domain (with multiple meshes).]{Front-right side of the structure within the computational domain (with multiple meshes). Each of the 16 meshes is 3~m (9.8~ft) by 10~m (33~ft) by 5~m (16.4~ft).}
\label{fig:mult_mesh}
\end{figure}

\begin{table}
\centering
\captionof{table}{Relevant fire model input parameters}\label{tab:mod_param}
\begin{tabular}{lll}
\toprule[1.5pt]
Parameter & Description & Reference  \\
\midrule
 Simulation Time                   &  5 min                                                        & -- \\[.25cm]
 Grid Cell Size                    &  10 cm                                                        & Section~\ref{mesh} \\ [.25cm]
 Ambient Temperature               &  25 $^{\circ}$C                                               & -- \\ [.1cm]
\multirow{4}{*}{Reaction: Wood}    &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$                     & \multirow{4}{*}{Section~\ref{fire}} \\
                                   &  CO Yield: 0.004 \; ${\mathrm{kg_{CO}}/\mathrm{kg_{wood}}}$   & \\
                                   &  Soot Yield: 0.015 \; ${\mathrm{kg_{C}}/\mathrm{kg_{wood}}}$  & \\
                                   &  Heat of Combustion: 16,400~kJ/kg                             & \\ [.25cm]
\multirow{2}{*}{Peak HRR}          &  Attic Fire: 9~MW                                             & \multirow{2}{*}{Section~\ref{fire}}\\ 
                                   &  Porch Fire: 3.3~MW                                           & \\ [.25cm]                     
\multirow{3}{*}{Material: Wood}    &  k: 0.12 \; \si{W/(m.K)})                                     & \multirow{3}{*}{Section~\ref{matl}} \\
                                   &  $\rho$: 510 \; kg/m$^3$                                      & \\
                                   &  c$_{p}$: 1.38 \; \si{kJ/(kg.K)}                              & \\[.25cm]
\multirow{3}{*}{Material: Gypsum}  &  k: 0.17 \; \si{W/(m.K)}                                      &  \multirow{3}{*}{Section~\ref{matl}}\\ 
                                   &  $\rho$: 930 \; kg/m$^3$                                      & \\ 
                                   &  c$_{p}$: 1.09 \; \si{kJ/(kg.K)}                              & \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=.80\textwidth]{../Figures/Chicago_Fire_HRR}
\caption[HRR versus time from FDS simulations of the West 50th St. fire.]{HRR versus time from FDS simulations of the West 50th St. fire. The first vertical line indicates when the back door failed, and the second vertical line indicates when the interior hallway door failed.}
\label{fig:hrr}
\end{figure}

In Fig.~(\ref{fig:hrr}), the solid line represents the baseline FDS simulation (this follows the noted door openings from the incident). The dashed line represents an alternative series of events in which the rear door does not get opened and the subsequent breach of the rear porch wall on the left side of the structure does not occur. The dash-dotted line represents the prescribed (or nominal) HRR that was input into FDS. As shown in Fig.~(\ref{fig:hrr}), the HRR output from FDS is less than the prescribed HRR because the fire in the structure becomes ventilation limited. The maximum HRR value is slightly less than 9~MW before the fire becomes ventilation limited and begins to decay. The HRR is reduced to 5~MW as the oxygen inside the structure is consumed. There is an insufficient amount of oxygen in the structure to combust all of the fuel, therefore combustion occurs locally at the vents where fresh air is available. Figure~(\ref{fig:smv_ext_fire}) shows flaming combustion through the failed window on the rear side of the structure and attic breach just after the second fire ignites at 40~s. After additional wall breaches occur at 45~s and 55~s (see Table~\ref{tab:vents}), the abrupt change in ventilation causes the HRR to increase to 9~MW before it decreases to approximately 6~MW. The HRR then increases towards a sustainable steady-state value.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=.675\textwidth]{../Figures/smv_exterior_fire}
\caption{Smokeview rendering of exterior combustion occurring on the rear side of the structure.}
\label{fig:smv_ext_fire}
\end{figure}

@@

As the fire grows in the attic and rear porch, the pressure rises. Figure~(\ref{fig:pres_159s}) shows the pressure conditions in the structure just before the interior door failed in the simulation. On the rear porch, the pressure rise ranges from a 4~Pa over pressure 1~m off the floor to greater than 10~Pa (0.00145~psi) at the ceiling. Gases at an elevated pressure will always move towards a region of lower pressure. The door serves as a boundary between the high and low pressure regions. Once the door fails open, the high pressure gases flow towards the low pressure regions in the house, hallway, kitchen, and front stairwell. The velocity vectors shown in Fig.~(\ref{fig:velo_175s}), 15~s after the interior door fails, confirm this flow direction within the structure.
